Parsing Postfix log files

### 2009-01-16 doc/logparser-other-parsers-review.tex
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid false positives. Improved the check for contractions and fixed any it found. Made running make somewhat quieter, to avoid clutter that conceals more useful     output. Improved the underfull/overfull check, a

At the start of this project ten Postfix log file parsers were tested, with
the hope of finding a suitable parser to build upon, rather than starting
from scratch.  There are not many Postfix log file parsers available ---
indeed it was quite difficult to find ten parsers to review for this
project --- and the functionality offered ranges from quite basic to much
more mature, depending on the needs of the author of the parser.  None of
those parsers were suitable for this project, so the decision was taken to
write a new parser.  The first parser reviewed is the only previously
published research in this area that the author is aware of; it parses
Postfix log files, but aim of the research is to show that presenting
extracted data in a more easily accessible format is useful to systems
administrators, rather than to improve anti-spam techniques..

The same ten parsers have been reviewed and compared to this project's
finished parser, to show how much effort would have been required to fulfil
the aims and requirements of this project.  It is important to compare and
contrast newly developed algorithms and parsers against those already
available, to accurately judge what improvements, if any, are delivered by
the newcomers.  Each of the reviewed parsers was tested with the
\numberOFlogFILES{} test log files described in \sectionref{parser
efficiency}.  \Tableref{Summary of parsers' features} summarises the
results.

There are some important differences between \parsername{} and the parsers
reviewed here:

@@

    \item Most of the reviewed parsers silently ignore log lines they
        cannot handle, whereas \parsername{} complains loudly about every
        single log line it fails to parse.  The exception is AWStats, which
        outputs the percentage of input lines it was unable to parse, but
        does not output the lines themselves.

    \item A minor difference is that most parsers do not handle compressed
        files; both \parsername{} and Splunk handle them transparently,
        without user intervention; Sawmill and Lire can be configured to
        support compressed files, but Sawmill exhibits a dramatic increase
        in parsing time when doing so.  Support for reading compressed log
        files is quite helpful, as it dramatically reduces the disk space
        required to store historical log files.

    \item Most of the reviewed parsers do not distinguish between different
        delivery attempt rejections, so they can not be used to determine
        the success rate of different anti-spam techniques.  The exception
        is Pflogsumm, which provides a summary of why delivery attempts
        were rejected.

@@

There only appears to be one prior published paper about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \gls{LMA} is quite
different from \parsername{}: it attempts to present correlated data from
log files in a form suitable for a systems administrator to search using
the myriad of standard Unix text processing utilities already available.
It produces a \gls{CSV} file and either a MySQL
(\url{http://www.mysql.com/}) or Berkeley DB
(\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
The decision to support both \gls{CSV} and Berkeley DB appears to have been
a serious limitation: both formats have limitations that will be explained
later.  Very little documentation is provided with \gls{LMA}, though some
documentation is available in~\cite{log-mail-analyser}.  Studying the
source code is informative, though this author had difficulty as the
authors of \gls{LMA} wrote in Italian.

\gls{CSV} is a very simple format where each record is stored in a single
line, with fields separated by a comma or other punctuation symbol.
Problems with \gls{CSV} files include the need to escape separators in the
data stored, providing multiple values for a field (e.g.\ multiple
recipients), and adding new fields.  There is no standard mechanism to
document the fields or the separator, unlike \gls{SQL} databases where
every database includes a schema naming the fields and the type of data
they store (integer, text, timestamp, etc.).  The \gls{CSV} record format
is not documented, but the output file contains a comment giving the
format:\newline{} \texttt{\# Timestamp|Nome Client|IP Client|IP
Server|From|To|Status|Size} \newline{}\gls{LMA} treats lines starting with
\texttt{\#} as comments, but not all \gls{CSV} parsers will.

Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
unlike \gls{SQL} databases that store arbitrary tuples.  In \gls{LMA}'s
main table the key is an integer referred to by secondary tables, and the
value is a \gls{CSV} line containing all of the data for that row.  The
secondary by-sender, by-recipient, by-date, and by-\gls{IP} tables use the
sender/recipient/date/\gls{IP} address as the key, and the value is a
\gls{CSV} list of integers referring to the main table.  This effectively
re-implements \gls{SQL} foreign keys, but without the functionality offered
by even the most basic of \gls{SQL} databases (joins, ordering, searches,
etc.).  It also requires custom code to search on some combination of the
above, though the authors of \gls{LMA} did provide some queries: IP-STORY,
FROM-STORY, DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears to be the
least useful of the three output formats: it does not provide the
functionality of a basic \gls{SQL} database, and unlike \gls{CSV} files it
can not be used with standard Unix text processing tools.

The schema used with the MySQL database is undocumented, but at least it is
possible to discover the schema with an existing \gls{SQL} database, unlike with
Berkeley DB\@; all \gls{SQL} databases embed the schema into the database
and provide commands for displaying it.  Berkeley DB does not embed a
schema, as there is neither requirement nor benefit; it only provides
\textbf{(key, value)} pairs, so any additional structuring of the data is
imposed by the application, thus the application must document this
structure.  MySQL support was not tested because there is no documentation
on the schema required.

Whether a MySQL database or Berkeley DB table is chosen in addition to the
\gls{CSV} output, \gls{LMA} stores the following data: time and date,
client hostname and \gls{IP} address, server \gls{IP} address, sender and
recipient addresses, \gls{SMTP} code, and size (for accepted mails only).
It is unclear which time and date is stored: start time, end time, or
delivery time?  Unlike \parsername{} it does not store the server hostname,
HELO hostname, queueid, start and end times, timestamps for each log line,
or message id (for accepted mails only).  Handling of multiple recipients,
\gls{SMTP} codes, or remote servers\footnote{A single mail may be sent to
multiple remote servers if it was addressed to recipients in different
domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained; experimental observation shows that multiple
records are added when there are multiple recipients (sadly the records are
not associated or linked in any way), and presumably the same approach is
taken when there are multiple destination servers.

@@

\gls{LMA} does not deal with any of the complications discussed in
\sectionref{complications}, except for correlating log lines by queueid;
not correlating log lines by pid means it cannot correlate most rejections.
It does not differentiate between different types of rejections, so it is
not suitable for the purposes of this project; the data about which
restriction caused the rejection is discarded, whereas the main goal of
this project is to retain that data to aid optimisation and evaluation of
anti-spam techniques.  \gls{LMA} fails to parse Postfix log files generated
on Solaris hosts because the fields automatically prepended to each log
line differ from those added on Linux hosts; log files from Solaris hosts
(and possibly other operating systems) thus require preprocessing before
parsing by \gls{LMA}.  Once the preprocessing has been performed on the
\numberOFlogFILES{} test log files \gls{LMA} parses the log files without
complaint, although it produced 32 entries in its output file for every
rejection in the input log file; it also missed some 40\% of delivered
mail.  Once these glaring deficiencies were discovered the author did not
waste any more time checking the results.

@@

    \item \gls{LMA} produces 32 output lines in its \gls{CSV} file for
        every single line it should have produced; the command above
        suppresses duplicate sequential lines.  DAILY-REJECT produces the
        correct number of output lines, probably because it uses the
        Berkeley DB files as input rather than the \gls{CSV} file.

@@

There are some differences between the output from DAILY-REJECT and the
\texttt{awk} command; the author did not spend substantial time attempting
to explain these differences.

@@

    \item The output from DAILY-REJECT is missing some records which are
        present in the \gls{CSV} file; this may be because it uses the
        Berkeley DB files instead, and there may be differences between the
        contents.

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any data, and it does not
extract any data that it does not require to produce its report, e.g.\ it
does not extract the HELO hostname, queueid, start and end times,
timestamps for each log line, or message id.  Both the parsing and
reporting are difficult to extend because it is a specialised tool, unlike
the easily extensible design of \parsername{}.  It does not correlate log
lines by queueid or \gls{pid}, and does not need to deal with the
complications encountered during this project (\sectionref{complications}).
Pflogsumm produces a useful report, and successfully parsed the
\numberOFlogFILES{} log files it was tested with.\footnote{The results it
reported were not verified in detail, but it did not report any errors, and
has a very good reputation amongst Postfix users.}  Pflogsumm has many
options to include or exclude certain sections of the report, all clearly
documented; by default it includes the following:

@@

Pflogsumm is the only reviewed parser which distinguishes between different
rejections.

@@

Sawmill is a general purpose commercial product that parses 818 log file
formats (as of 2008/11/23) and produces reports from the extracted data.
Its data extraction facilities (described later) are too limited to save
sufficient data for the purposes of this project: although it can extract
three different sets of data from Postfix log files, they are not
interlinked in any way.  The documentation does not suggest that any
attempt is made to correlate log lines by either queueid or pid or to deal
with the other difficulties documented in \sectionref{complications}.

@@

    \item \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{}
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin \gls{IP} address, and virus.
        It also counts the number of and total size of all mails delivered.
        The fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
        \texttt{virus} are not explained in the documentation:
        \texttt{server} is probably the hostname or \gls{IP} address of the
        server the mail is delivered to; \texttt{relay} might be the
        delivery method: \gls{SMTP}, local delivery, or \gls{LMTP};
        \texttt{uid} might be the uid of the user submitting mail locally.
        Postfix does not perform any form of virus checking (though it has
        many options for cooperating with an external virus scanner), so
        the \texttt{virus} field is a mystery.

@@

The source code is available in an obfuscated form, and the product is
quite expensive, requiring a \euros{100} + VAT licence per report
(discounts are available when buying multiple licences); in contrast
\parsername{} is free to use and the code is freely available.  Sawmill is
supplied with thorough and well written documentation; everything the
author searched for was documented, except the MySQL database schema and
some details of the data extracted by the parser.  A commercial version of
MySQL is required due to MySQL licensing restrictions, but Sawmill's
documentation explains why and includes instructions on how to compile
Sawmill so that it can use a non-commercial version of MySQL (this was not
attempted during the review process).

Sawmill's web interface supports searching on any combination of the fields
it extracts, and searches performed using the web interface produced
accurate results.  The interface for searching is neither as simple to use
nor as informative as the interface provided by Splunk.  The administrative
interface is much easier to use than Splunk's: it took only five minutes to
start parsing a whole directory of log files.

@@

\parserblurb{%
    Splunk is IT Search.  Search and navigate IT data from applications,
    servers and network devices in real-time. Logs, configurations,
    messages, traps and alerts, scripts, code, metrics and more. If a
    machine can generate it --- Splunk can eat it. It's easy to download
    and use and it's very powerful.
}
{http://www.splunk.com/}
{2008/11/23}

XXX THE FIRST PARAGRAPH NEEDS TO BE REWRITTEN\@; PICK POINTS, THEN COMPARE
AND CONTRAST\@.

Splunk aims to index all of an organisation's log files, providing a
centralised view capable of searching and correlating diverse log sources.
The web interface supports complicated searches, providing statistics and
graphs in real time, a facility not provided by \parsername{} (report
generation has been deferred to a subsequent program).  Searches can be
based on the fields extracted by Splunk or the full text of the log line.
Splunk allows quite complicated searches, but does not make the raw data
available in an accessible form.  Saved searches can be run periodically
and the results emailed to a recipient or sent to an external program for
further processing (though maybe without the graphs and detailed
statistics); the author was unable to save searches, though that may have
been due to limitations in the free version.  The database is not available
for use by external programs, whereas \parsername{} provides the database
and leaves it to the user to utilise it without limit or restriction.  The
interface is optimised for interactive use rather than automated queries
and it does not appear to be possible to write independent tools to utilise
the Splunk database;.  Some additional Postfix reports are supposedly
available at \url{http://www.splunkbase.com/}, but the author was unable to
find any Postfix reports, or indeed for any other log file types: every
category was empty, even those that the web site claimed had many reports
available.  Many types of graphs can be generated, though most are
variations of a bar or pie chart, except bubble and heatmap graphs.  It is
easy to drill down through the graphs to extract a portion of the data
(e.g.\ select the hour with the largest number of events, then select a
particular host, and finally a specific address), though it is not possible
to search on partial words.  All searches performed using the indexed data
returned reasonable results.

The web interface is quite attractive and simple to use when searching, but
as an administrator it seems unnecessarily difficult to perform simple
tasks.  When testing Splunk it took roughly 30 minutes to figure out how to
add a single log file to be indexed so that it could be searched, with the
downside that the log file was copied into a spool directory before
indexing, doubling the disk space usage.  The next test was to index all
the log files in a particular directory, but after three hours, numerous
futile attempts, and reading all the available documentation, the author
admitted defeat.  Using the \gls{CLI} rather than the web interface was
partially successful: the command \newline{} \tab{} \texttt{splunk find
logs }\textit{log-directory\/}\newline{} added 40 of the
\numberOFlogFILES{} log files to the queue for indexing.  Further attempts
enqueued the same 40 log files, without explaining why the others were
excluded.\footnote{The log files appear to have been indexed once only;
presumably Splunk keeps track of the files it has indexed and discards
requests to index files for a second time.  This may or may not be a useful
feature for \parsername{}.} There did not appear to be an option to ensure
the log files would be processed in the order they were created, though
this may be neither necessary nor beneficial with Splunk.  Subsequently the
author was successful in adding a single file at a time using the
\gls{CLI}:\newline{} \tab{} \texttt{splunk add tail
}\textit{filename\/}\newline{} A simple loop was then enough to add all the
desired log files.  Splunk will periodically check all indexed files for
updates unless they are manually removed from its list; this may or may not
be useful behaviour.  Splunk did not appear to have any difficulty in
indexing the log files, once they had been successfully added to its queue.
\parsername{} parses the logs it is instructed to parse, in the order
given; periodic parsing of logs is a task an administrator can easily
achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.

@@

Splunk supports reading compressed log files without any configuration by
the user.  The free version of Splunk limits the volume of data indexed per
day to 500MB, though a trial Enterprise licence is available that allows
indexing of up to 5GB of data per day.  In 2007 the cheapest licenced
version cost \$5000 plus \$1000 support, and limited the volume of data
indexed per day to 500MB\@.  Prices were removed from the Splunk website
during 2008; now Splunk's sales team must be contacted for a quote.
Typical log file sizes for a small scale mail server are given in
\sectionref{parser efficiency}.

@@

Splunk is a generic tool, so it lacks any Postfix specific support over and
above extracting the \texttt{key=value} fields from a log line; most
importantly it makes no attempt to correlate log lines by queueid or
\gls{pid}, or to handle any of the other myriad complications discussed in
\sectionref{complications}.  Its source code is unavailable, so it could
not be used as a base for this project, even if it fulfilled all other
requirements.

@@

Isoqlog's report misses most of the information gathered by \parsername{}:
the data extracted is limited to the number of mails sent by each sender,
and it only reports on senders from the domains listed in its configuration
file, making it impossible to produce complete reports.  It ignores all log
lines except those with today's date, so it is impossible to analyse
historical log files, and testing with the \numberOFlogFILES{} test log
files was pointless.  It does maintain a record of data previously
extracted, which the newly extracted data is merged into; the format of the
data store is undocumented.  Very little documentation is provided with
Isoqlog, little more than installation instructions.  It does not utilise
rejection log lines in any way, so is unsuitable for the purposes of this
project.  Its parsing is completely inextensible, indeed is almost
incomprehensible, relying on \texttt{scanf(3)}, unexplained fixed offsets,
and low level string manipulation; it is the opposite end of the spectrum
to \parsernames{} parsing.  It does not handle any of the complications
discussed in \sectionref{complications}, does not gather the breadth of
data required for this project, and ignores the majority of log lines
produced by Postfix.

@@

\parserblurb{%
    AWStats is a free powerful and featureful tool that generates advanced
    web, streaming, ftp or mail server statistics, graphically. This log
    analyzer works as a CGI or from command line and shows you all possible
    information your log contains, in few graphical web pages. It uses a
    partial information file to be able to process large log files, often
    and quickly. It can analyze log files from all major server tools like
    Apache log files (NCSA combined/XLF/ELF log format or common/CLF log
    format), WebStar, IIS (W3C log format) and a lot of other web, proxy,
    wap, streaming servers, mail servers and some ftp servers.
}
{http://awstats.sourceforge.net/}
\url{http://awstats.sourceforge.net/awstats.mail.html} \newline{}
\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} \newline{}
(Last checked 2009/01/11.)

AWStats can produce simple graphs from many different services' log files,
but supporting numerous log files formats without special purpose code
limits its functionality.  The data it can extract from an \gls{MTA} log
file is limited in comparison to \parsername{}: time2, email, email\_r,
host, host\_r, method, url, code, and bytesd.  There is no explanation for
any of those fields in the documentation (in contrast \parsername{}
documents everything it extracts), so the author could not understand what
data is extracted, nor determine what data is missing in comparison to
\parsername{}.  AWStats coerces Postfix log files into Apache\footnote{The
Apache web server is the most popular HTTP server in use over the past 10
years; more information is available at \url{http://httpd.apache.org/}.}
format log files, for analysis by AWStats' HTTP log file parser.  The
converting parser only deals with a small portion of the log lines
generated by Postfix, silently skipping those it cannot deal with, and does
not distinguish between different types of rejection; extending it to parse
all log lines would be at least as much work as writing a new parser.  It
does correlate log lines by queueid (not by pid), but it does not deal with
any of the other complications described in \sectionref{complications}.
AWStats supports saving data, but the format of the data store is not
documented.  It also supports reading compressed log files, but that
functionality was not tested.

>>>>

At the start of this project ten Postfix log file parsers were tested, with
the hope of finding a suitable parser to build upon, rather than starting
from scratch.  It was quite difficult to find ten parsers to review for
this project, and the functionality offered by those parsers ranges from
quite basic to much more mature, depending on the needs of the author of
the parser.  None of those parsers were suitable for this project, so the
decision was taken to write a new parser.  The first parser reviewed is the
only previously published research in this area that the author is aware
of; it parses Postfix log files, but aim of the research is to show that
presenting extracted data in a more easily accessible format is useful to
systems administrators, rather than to improve anti-spam techniques.

The same ten parsers have been reviewed and compared with \parsername{},
this project's finished parser, to show how much effort would have been
required to fulfil the aims and requirements of this project.  It is
important to compare and contrast newly developed algorithms and parsers
against those already available, to accurately judge what improvements, if
any, are delivered by the newcomers.  Each of the reviewed parsers was
tested with the \numberOFlogFILES{} test log files described in
\sectionref{parser efficiency}.  \Tableref{Summary of parsers' features}
summarises the results.

Some important differences exist between \parsername{} and most or all of
the parsers reviewed here:

@@

    \item Most of the reviewed parsers silently ignore log lines they
        cannot handle, whereas \parsername{} complains loudly about every
        single log line it fails to parse.  The exception is AWStats, which
        outputs the percentage of log lines it was unable to parse, but
        does not output the log lines themselves.

    \item A minor difference is that most parsers do not handle compressed
        log files; both \parsername{} and Splunk handle them transparently,
        without user intervention; Sawmill and Lire can be configured to
        support compressed log files, but Sawmill exhibits a dramatic
        increase in parsing time when doing so.  Support for reading
        compressed log files is quite helpful, as it dramatically reduces
        the disk space required to store historical log files.

    \item Most of the reviewed parsers do not distinguish between different
        delivery attempt rejections, so they cannot be used to determine
        the success rate of different anti-spam techniques.  The exception
        is Pflogsumm, which provides a summary of why delivery attempts
        were rejected.

@@

There only appears to be one prior published paper about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \gls{LMA} is quite
different from \parsername{}: it attempts to present correlated data from
log files in a form suitable for a systems administrator to search using
the myriad of standard Unix text processing utilities already available.
It produces a \gls{CSV} file and either a MySQL
(\url{http://www.mysql.com/}) or Berkeley DB
(\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
The decision to support both \gls{CSV} and Berkeley DB appears to have been
a serious limitation: both formats have limitations that will be explained
later.  Hardly any documentation is provided with \gls{LMA}, though some
documentation is available in~\cite{log-mail-analyser}.  Studying the
source code is informative, though this author had difficulty as the
authors of \gls{LMA} wrote in Italian.

\gls{CSV} is a very simple format where each record is stored in a single
line, with fields separated by a comma or other punctuation symbol.
Problems with \gls{CSV} files include the need to escape separators in the
data stored, providing multiple values for a field (e.g.\ multiple
recipients), and adding new fields.  \gls{CSV} files do not have a standard
mechanism to document the fields or the separator, unlike \gls{SQL}
databases where every database includes a schema naming the fields and the
type of data they store (e.g.\ integer, text, timestamp).  The \gls{CSV}
record format is not documented, but the output file contains a comment
giving the format:\newline{} \texttt{\# Timestamp|Nome Client|IP Client|IP
Server|From|To|Status|Size} \newline{}\gls{LMA} treats \gls{CSV} lines
starting with \texttt{\#} as comments, but not all \gls{CSV} parsers will.

Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
unlike \gls{SQL} databases that store arbitrary tuples.  In \gls{LMA}'s
main table the key is an integer referred to by secondary tables, and the
value is a \gls{CSV} line containing all of the data for that row.  The
secondary by-sender, by-recipient, by-date, and by-\gls{IP} tables use the
sender/recipient/date/\gls{IP} address as the key, and the value is a
\gls{CSV} list of integers referring to the main table.  This effectively
re-implements \gls{SQL} foreign keys, but without the functionality offered
by even the most basic of \gls{SQL} databases (e.g.\ joins, ordering,
searches).  It also requires custom code to search on some combination of
the above, though the authors of \gls{LMA} did provide some queries:
IP-STORY, FROM-STORY, DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears
to be the least useful of the three output formats: it does not provide the
functionality of a basic \gls{SQL} database, and unlike \gls{CSV} files it
cannot be used with standard Unix text processing tools.

The schema used with the MySQL database is undocumented, but at least it is
possible to discover the schema with an existing \gls{SQL} database, unlike
with Berkeley DB\@; all \gls{SQL} databases embed the schema into the
database and provide commands for displaying it.  Berkeley DB does not
embed a schema in its files, because there is neither requirement nor
benefit; it only provides \textbf{(key, value)} pairs, so any additional
structuring of the data is imposed by the application, thus the application
must document this structure.  MySQL support was not tested because the
schema required is not documented.

Whether a MySQL database or Berkeley DB table is chosen in addition to the
\gls{CSV} output, \gls{LMA} stores the following data: time and date,
client hostname and \gls{IP} address, server \gls{IP} address, sender and
recipient addresses, \gls{SMTP} code, and size (for accepted mails only).
It is unclear what time and date is stored: start time, end time, or
delivery time?  Unlike \parsername{} it does not store the server hostname,
HELO hostname, queueid, start and end times, timestamps for each log line,
or message id (for accepted mails only).  Handling of multiple recipients,
\gls{SMTP} codes, or remote servers\footnote{A single mail may be sent to
multiple remote servers if it was addressed to recipients in different
domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained; experimental observation shows that multiple
records are added when a mail had multiple recipients (sadly the records are
not associated or linked in any way), and presumably the same approach is
taken when there were multiple destination servers.

@@

\gls{LMA} does not deal with any of the complications discussed in
\sectionref{complications}, except for correlating log lines by queueid;
not correlating log lines by pid means it cannot correlate most rejections.
It does not differentiate between different types of rejections, so it is
not suitable for the purposes of this project; the data about which
restriction caused the rejection is discarded, whereas the main goal of
this project is to retain that data to aid optimisation and evaluation of
anti-spam techniques.  \gls{LMA} fails to parse Postfix log files generated
on Solaris hosts because the fields automatically prepended to each log
line differ from those added on Linux hosts; log files from Solaris hosts
(and possibly other operating systems) thus require preprocessing before
parsing by \gls{LMA}.  The \numberOFlogFILES{} preprocessed test log files
were parsed without complaint by \gls{LMA}, although it produced 32 entries
in the output \gls{CSV} file for every rejection in the input log file; it
also missed some 40\% of delivered mail.  Once these glaring deficiencies
were discovered the author did not waste any more time checking the
results.

@@

    \item \gls{LMA} produces 32 lines in its \gls{CSV} file for every
        single line it should have produced; the command above suppresses
        duplicate sequential lines.  DAILY-REJECT produces the correct
        number of output lines, probably because it uses the Berkeley DB
        files as input rather than the \gls{CSV} file.

@@

The output from DAILY-REJECT and the \texttt{awk} command is not exactly
the same; the author did not spend substantial time attempting to explain
these differences.

@@

    \item The output from DAILY-REJECT is missing some records that are
        present in the \gls{CSV} file; this may be because it uses the
        Berkeley DB files instead, and there may be differences between the
        contents.

@@

Pflogsumm produces a report designed for troubleshooting rather than
in-depth analysis.  It does not support saving any data, and it does not
extract any data that it does not require to produce its report, e.g.\ it
does not extract the HELO hostname, queueid, start and end times,
timestamps for each log line, or message id.  Both the parsing and
reporting are difficult to extend because it is a specialised tool, unlike
the easily extensible design of \parsername{}.  It does not correlate log
lines by queueid or \gls{pid}, and does not need to deal with the
complications encountered during this project (\sectionref{complications}).
Pflogsumm produces a useful report, and successfully parsed the
\numberOFlogFILES{} log files it was tested with.\footnote{The results it
reported were not verified in detail, but it did not report any errors, and
has an excellent reputation amongst Postfix users.}  Pflogsumm has many
options to include or exclude certain sections of the report, all clearly
documented; by default it includes the following:

@@

Pflogsumm is the only reviewed parser that distinguishes between different
rejections.

@@

Sawmill is a general purpose commercial product that parses 818 log file
formats (as of 2008/11/23) and produces reports from the extracted data.
Its data extraction facilities (described later) are too limited to save
enough data for the purposes of this project: although it can extract three
different sets of data from Postfix log files, they are not interlinked in
any way.  The documentation does not suggest that Sawmill correlates log
lines by either queueid or pid or deals with the other difficulties
documented in \sectionref{complications}.

@@

    \item \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{}
        Fields extracted: from, to, server, UID, relay, status, number of
        recipients, origin hostname, origin \gls{IP} address, and virus.
        It also counts the number of and total size of all mails delivered.
        The fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
        \texttt{virus} are not explained in the documentation:
        \texttt{server} is probably the hostname or \gls{IP} address of the
        server the mail is delivered to; \texttt{relay} might be the
        delivery method: \gls{SMTP}, local delivery, or \gls{LMTP};
        \texttt{uid} might be the uid of the user submitting mail locally.
        Postfix does not do any form of virus checking (though it has many
        options for cooperating with an external virus scanner), so the
        \texttt{virus} field is a mystery.

@@

The source code is available in an obfuscated form, and the product is
quite expensive, requiring a \euros{100} + VAT licence per report
(discounts are available when buying multiple licences); in contrast
\parsername{} is free to use and the code is freely available.  Sawmill is
supplied with thorough and well written documentation; everything the
author searched for was documented, except the MySQL database schema and
some details of the data extracted by the parser.  A commercial version of
MySQL is required because of MySQL licensing restrictions, but Sawmill's
documentation explains why and includes instructions on how to compile
Sawmill so that it can use a non-commercial version of MySQL (this was not
attempted during the review process).

Sawmill's web interface supports searching on any combination of the fields
it extracts, and all searches produced accurate results.  The interface for
searching is neither as simple to use nor as informative as the interface
provided by Splunk.  The administrative interface is much easier to use
than Splunk's: it took only five minutes to start parsing a whole directory
of log files.

@@

\parserblurb{%
    Splunk is IT Search.  Search and navigate IT data from applications,
    servers and network devices in real-time. Logs, configurations,
    messages, traps and alerts, scripts, code, metrics and more. If a
    machine can generate it --- Splunk can eat it. It\empty{}'s easy to
    download and use and it\empty{}'s very powerful.
}
{http://www.splunk.com/}
{2008/11/23}

Splunk aims to index all of an organisation's log files, providing a
centralised view capable of searching and correlating diverse log sources.
The web interface provides search functionality, generating statistics and
graphs in real time, a facility not provided by \parsername{}.  Splunk
allows quite complicated searches, based on the fields extracted by Splunk
(described later) or the full text of the log line, though it is not
possible to search on partial words.  Search results are not available for
use with other tools.  Searches can be saved for reuse; saved searches can
be run periodically and the results emailed to a recipient or sent to an
external program for further processing.  The author was unable to save
searches, though that may have been because of limitations in the free
version.  The data store is not available for use by external programs,
whereas \parsername{} provides the database and leaves it to the user to
utilise it without limit or restriction.  The web interface is optimised
for interactive use rather than automated queries, and it does not appear
to be possible to write independent tools to utilise the Splunk database.
Some additional Postfix reports are supposedly available on
\url{http://www.splunkbase.com/}, but the author was unable to find any
Postfix reports, or indeed reports for any other log file types: every
category was empty, even those that the web site claimed had numerous
reports available.  Many types of graphs can be generated, though most are
variations of a bar or pie chart, except the bubble and heatmap graphs.  It
is easy to drill down through the graphs to select a portion of the data
(e.g.\ select the hour with the largest number of events, then select a
particular host, and finally a specific sender address).  All searches
performed using the indexed data returned reasonable results.

The web interface is quite attractive and simple to use when searching, but
as an administrator it seems unnecessarily difficult to perform simple
tasks.  When testing Splunk it took roughly 30 minutes to figure out how to
add a single log file to be indexed so that it could be searched, with the
downside that the log file was copied into a spool directory before
indexing, doubling the disk space usage.  The next test was to index all
the log files in a particular directory, but after three hours, numerous
futile attempts, and reading all the available documentation, the author
admitted defeat.  Using the \gls{CLI} rather than the web interface was
partially successful: the command \newline{} \tab{} \texttt{splunk find
logs }\textit{log-directory\/}\newline{} added 40 of the
\numberOFlogFILES{} log files to the queue for indexing.  Further attempts
enqueued the same 40 log files, without explaining why the others were
excluded.\footnote{The log files appear to have been indexed once only;
presumably Splunk keeps track of the log files it has indexed and discards
requests to index log files for a second time.  This may or may not be a
useful feature for \parsername{}.} There did not appear to be an option to
ensure the log files would be processed in the order they were created,
though this may be neither necessary nor beneficial with Splunk.
Subsequently the author was successful in adding a single file at a time
using the \gls{CLI}:\newline{} \tab{} \texttt{splunk add tail
}\textit{filename\/}\newline{} A simple loop was then enough to add all the
desired log files.  Splunk will periodically check all indexed log files
for updates unless they are manually removed from its list; this may or may
not be useful behaviour.  Splunk did not appear to have any difficulty in
indexing the log files, once they had been successfully added to its queue.
\parsername{} parses the logs it is instructed to parse, in the order
given; periodic parsing of logs is a task an administrator can easily
achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.

@@

Splunk supports reading compressed log files without any configuration by
the user.  The free version of Splunk limits the volume of data indexed per
day to 500MB, though a trial Enterprise licence is available that allows
indexing of up to 5GB of data per day.  In 2007 the cheapest licenced
version cost \$5000 plus \$1000 support, and limited the volume of data
indexed per day to 500MB\@.  Prices were removed from the Splunk website
during 2008; now Splunk's sales team must be asked for a quote.  Typical
log file sizes for a small scale mail server are given in
\sectionref{parser efficiency}.

@@

Splunk is a generic tool, so it lacks any Postfix specific support over and
above extracting the \texttt{key=value} fields from a log line; it makes no
attempt to correlate log lines by queueid or \gls{pid}, or to handle any of
the other myriad complications discussed in \sectionref{complications}.
Its source code is unavailable, so it could not be used as a base for this
project, even if it fulfilled all other requirements.

@@

Isoqlog's report misses most of the information gathered by \parsername{}:
the data extracted is limited to the number of mails sent by each sender,
and it only reports on senders from the domains listed in its configuration
file, making it impossible to produce complete reports.  It ignores all log
lines except those with today's date, so it is impossible to analyse
historical log files, and testing with the \numberOFlogFILES{} test log
files was pointless.  It does maintain a record of data previously
extracted and the newly extracted data is merged into it; the format of the
data store is undocumented.  Almost no documentation is provided with
Isoqlog, little more than installation instructions.  It does not utilise
rejection log lines in any way, so is unsuitable for the purposes of this
project.  Its parsing is completely inextensible, indeed is almost
incomprehensible, relying on \texttt{scanf(3)}, unexplained fixed offsets,
and low level string manipulation; it is the opposite end of the spectrum
to \parsernames{} parsing.  It does not handle any of the complications
discussed in \sectionref{complications}, does not gather the breadth of
data required for this project, and ignores most of the log lines produced
by Postfix.

@@

\parserblurb{%
    AWStats is a free powerful and featureful tool that generates advanced
    web, streaming, ftp or mail server statistics, graphically. This log
    analyzer works as a CGI or from command line and shows you all possible
    information your log contains, in few graphical web pages. It uses a
    partial information file to be able to process large log files, often
    and quickly. It can analyze log files from all major server tools like
    Apache log files (NCSA combined/XLF/ELF log format or common/CLF log
    format), WebStar, IIS (W3C log format) and a lot of other web, proxy,
    wap, streaming servers, mail servers and some ftp servers.
}
{http://awstats.sourceforge.net/awstats.mail.html}
{2009/01/11}

AWStats can produce simple graphs from many different services' log files,
but supporting numerous log files formats without special purpose code
limits its functionality.  The data it can extract from an \gls{MTA} log
file is limited in comparison to \parsername{}: time2, email, email\_r,
host, host\_r, method, url, code, and bytesd.  No explanation for any of
those fields is provided in the documentation at
\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} (in
contrast \parsername{} fully documents everything it extracts), so the
author could not understand what data is extracted, or determine what data
is missing in comparison to \parsername{}.  AWStats coerces Postfix log
files into Apache\footnote{The Apache web server is the most popular HTTP
server in use over the past 10 years; more information is available at
\url{http://httpd.apache.org/}.} format log files, for analysis by AWStats'
HTTP log file parser.  The converting parser only deals with a small
portion of the log lines generated by Postfix, silently skipping those it
cannot deal with, and does not distinguish between different types of
rejection; extending it to parse all log lines would be at least as much
work as writing a new parser.  It does correlate log lines by queueid (not
by pid), but it does not deal with any of the other complications described
in \sectionref{complications}.  AWStats supports saving data, but the
format of the data store is not documented.  It also supports reading
compressed log files, but that functionality was not tested.

When tested with the \numberOFlogFILES{} test log files AWStats' reported
that it parsed 9,240,075 (88.70\%) of 10,416,129 log lines, skipping
1,176,050 (11.29\%) corrupt log lines; the \numberOFlogFILES{} log files
contain \numberOFlogLINES{} log lines, so AWStats parsed only 15.21\% of
the log lines, declared 1.93\% were corrupt, and ignored the remaining
82.85\%.
